### 2014-06-07 paper/chemokine_supplement_revised.tex
R4 in progress

We estimate chemokine production rates, `r', by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} \drew{Eq. 1} by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\label{eq:dde}
\end{equation}
\vspace{.05in}
}

$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table 1 summarizes population and parameter values and descriptions.  Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table 3).

{\footnotesize
\begin{equation*}
\label{EqS1}
\begin{aligned}
\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
\dot{T} &= \rho T -\beta TV \\
\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
\dot{V} &= pI - \beta TV - \gamma (t) V \\
\gamma (t) &= \left\{ \begin{array}{rcl}
	1/\mbox{day} & \mbox{,}  & t < 5  \\
	3/\mbox{day} & \mbox{,} & t \geq 5  
	\end{array}\right. \\
\end{aligned}
\tag{S1}
\end{equation*}
}
\vspace{.05in}

@@

\subsection*{T cell sensitivity to chemokine}

@@

@@

@@

\begin{figure}[!ht]
\begin{center}
% \includegraphics[width=\textwidth]{Figure_S1}
 \end{center}
\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES. Total number of incubating, secreting and apoptotic cells are plotted for each infection.  The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it. } 
 \label{fig:sensitivity}
\end{figure}

\begin{figure}[ht!]
\begin{center}
%	\includegraphics[width=\textwidth]{Figure_S2}
	\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.  B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
	\label{fig:chemokine}
\end{center}
\end{figure}

@@

>>>>

@@

We estimate chemokine production rates, `r', by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\label{eq:dde}
\end{equation}
\vspace{.05in}
}

$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S1 summarizes population and parameter values and descriptions.  Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table 1).

%\subsection*{Chemokine Secretion}

%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).  The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..

%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref. \cite{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).  Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There was no positive correlation between viral production and induced chemokine production across the three strains.

{\footnotesize
\begin{equation*}
\label{EqS1}
\begin{aligned}
\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
\dot{T} &= \rho T -\beta TV \\
\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
\dot{V} &= pI - \beta TV - \gamma (t) V \\
\gamma (t) &= \left\{ \begin{array}{rcl}
	1/\mbox{day} & \mbox{,}  & t < 5  \\
	3/\mbox{day} & \mbox{,} & t \geq 5  
	\end{array}\right. \\
\end{aligned}
\tag{S1}
\end{equation*}
}
\vspace{0.5in}

\subsection{T cell sensitivity to chemokine}

@@

\subsection{General sensitivity analysis}

Model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Table~\ref{tab:parameters}).  Due to the difficulty of estimating critical, unmeasurable values, we performed a sensitivity analysis for 16 parameters in order to test whether variation over an order of magnitude or more altered the infection outcome.  Individual parameters were varied over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.  Each parameter was varied over all three influenza strains, creating three sets of sensitivity plots (Figures~\ref{fig:asensitivity}-\ref{fig:psensitivity}).

We then categorize the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection.  When evaluating the results, we consider model runs that asymptotically converge to be qualitatively stable and model runs that do not converge to be qualitatively divergent.

\subsubsection{Stable Parameters}

The parameters in the first grouping did not affect the outcome of the infection unless they were adjusted to values outside the realm of possibility.  Specifically, each parameter in this group seems to lie within a range that does not significantly affect the asymptotic behavior of the model.  Of interest, this group can be mostly split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).  Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.  The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal effects.

The stability of these parameters makes sense in the context of our model.  Chemokine exists in our model to provide a chemical gradient that T cells may follow to the focus of infection.  The total quantity of the chemokine in the lung does not have a strong effect on the location and size of the gradient.  Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells may follow.

Similarly, T cells affect the model by clearing cells expressing virus.  As discussed in the main paper, the chemokine gradient creates areas of maximal concentration that attract all the T cells inside its basin of attraction.  Thus, most T cells are attracted to the same areas of the infection and overlap considerably.  Increasing T cell numbers and efficiency will not help clear the infection beyond a certain point.

While we have claimed that these parameters are qualitatively stable within a certain range, many of the parameters were set to values that did lead to a divergent model behavior.  We have deemed these deviations acceptable on an individual basis as described below.

Apoptosis time diverges slightly in the sH1N1 strain.  As stated earlier, its inclusion in this group is already suspect and its effects are described in more detail in the results section of the main paper.

Varying the chemokine decay rate results in stable model behavior for values one order of magnitude larger and smaller than the default.  A value two orders of magnitude smaller results in divergent behavior for both the aH5N1 strain and the sH1N1 strain and a value two orders of magnitude larger results in divergent behavior for only the sH1N1 strain.  This maximum decay rate used for all three strains corresponds to an implausible 18 second half-life.  The minimal value corresponds to a similarly implausible 50 hour half-life.  The fact that an extremely low decay rate can hinder clearance is interesting.  In this case, lack of decay allows the chemokine to diffuse homogeneously across the entire infection, removing the concentration gradient required by T cells to find the active areas of the infection.  This confirms that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.  Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection.

The chemokine diffusion rate only diverges in the aH5N1 strain for the highest value.  If anything, our estimate for the diffusion rate is already large as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.  Thus the diffusion rate we used based on Stokes-Einstein and water viscosity was likely within a physiologic range.

Chemokine secretion values differ between strains.  It is important to node that aH5N1 and sH1N1 show a threshold at the same concentration: near 1e-6 $pg/s\cdot cell$.  While this may seem contradictory, it is actually an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.  Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of divergence.

T cell circulation times were tested over a very large range and only diverge at the very edge of that range.  Because vascular and lymph circulation times are unknown, parameter values of 1 minute and 3 minutes in mice may be plausible and warrant further study.

The T cell kill rate only diverges on the lower end in sH1N1.  The extreme value corresponds to a T cell needing 100 minutes to induce the apoptosis of a single infected cell and is not biologically reasonable.  The intermediate value corresponds to a time of 33.3 minutes and is also unlikely.

T cell movement in tissue has been observed \cite{Egen2011}.  Thus, we consider the extreme values biologically implausible.

T cell decay parameters allow the model to diverge only in the most extreme cases.  Neither of these values are reasonable and the parameters show stable behavior otherwise.

\subsubsection{Difference in Peak Only}

Two parameters, viral incubation time and viral expression time, do seem to affect the model's behavior up until the introduction of the T cell response, followed by convergence of infected cell numbers back to the same levels.  This effect occurs because the T cells short circuit the lifespan of infected cells.  Once T cells arrive, virus secreting cells no longer survive for their full lifespan.  Rather, the secretion time becomes limited by the apoptosis time parameter.  Because these terms refer to how and when infected cells release the virus particles, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response.

\subsubsection{Unstable Parameters}

This leaves five parameters shown that appear to directly affect the result of the infection: IgM strength, infection sensitivity, T cell secretion rate, viral decay, and viral diffusion.  By comparing the three strains of influenza, we also know that the viral secretion rate affects the result.  Of interest, nearly all of these parameters are directly related to the behavior of the virus.  Only the T cell secretion rate and IgM parameters are not.

IgM strength is not inserted until day 4 of the simulation, but once it does it directly affects the decay rate of the virus, and does so for the rest of the simulation.  The higher the strength of this parameter, the less virus there is in the system.  Because the true effective strength of IgM is unknown in the context of this model, it is not a target of our investigation.

Viral decay has the exact same effect as the IgM strength parameter.  Its value directly determines how much virus remains in the system over the course of the infection.

Infection sensitivity describes the ability of the virus to infect healthy cells, yet its strength is directly related to how much virus is present in the area of the healthy cells.  Thus, increasing its value by any factor is equivalent to increasing the amount of virus by the same factor.  Thus, it behaves similarly to the IgM and viral decay parameters.

Shown elsewhere, the viral secretion rate has a strong effect on the outcome of the infection.  Similar to the previously discussed parameters, its value directly affects how much virus there is in the system.  Thus, it's affect is similar to the previous parameters.

Viral diffusion does not change how much virus there is in the system.  Rather, it determines how fast the virus may spread across the alveoli.  While its mechanism is different, its effect may be even stronger than the previous parameters.  By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may outpace the body's ability to generate a chemical gradient.  Thus T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection.

Finally, the T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.  Thus it is reasonable to assume that there is a threshold of T cell secretion, beyond which the dynamics of the infection do not change.  This is consistent with our observations of T cell clumping in areas of high chemokine concentration.  Increasing the number of T cells in the system does not seem to help beyond a certain point because the T cells overlap in space and become redundant.

@@

\section{Discussion}

\subsection{Chemokine directed T cell search}

The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.  Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.  Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.  Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.  The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.

A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.  The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.  For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.  This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.

Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.

@@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figure_S1}
 \end{center}
\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES. Total number of incubating, secreting and apoptotic cells are plotted for each infection.  The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it. } 
 \label{fig:sensitivity}
\end{figure}

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=\textwidth]{Figure_S2}
	\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.  B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
	\label{fig:chemokine}
\end{center}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{|  l  l  l  |}
  \hline
  Population & Description & Initial Value\\
  \hline
  $T$ & Healthy target epithelial cells & 1,000,000 \\
  $I_1$ & Virus-incubating cells & 0 \\
  $I_2$ & Virus-secreting cells & 0 \\
  $V$ & Free virus particles & 10,000 \\
  $F$ & Interferon quantity & 0 \\
  $C$ & Chemokine quantity (ng/mL) & 0 \\
  \hline
  \hline
  Parameter & Description & Value \\
  \hline
  $\beta$ & Viral infection rates &  \\
    & \hspace{2em} Avian & 5.3e-7 (PFU$\cdot$h)$^{-1}$ \\
    & \hspace{2em} Seasonal & 6.1e-7 (PFU$\cdot$h)$^{-1}$ \\
    & \hspace{2em} Pandemic &  2.7e-6 (PFU$\cdot$h)$^{-1}$\\
  $p$ & Viral production rates & \\
    & \hspace{2em} Avian & 0.20 (PFU/h) \\
    & \hspace{2em} Seasonal & 1.4 (PFU/h) \\
    & \hspace{2em} Pandemic & 18.3 (PFU/h) \\
  $e$ & Interferon strengths & \\
    & \hspace{2em} Avian & 1.0e-8 \\
    & \hspace{2em} Seasonal & 1.6e-6 \\
    & \hspace{2em} Pandemic & 3.4e-3 \\
  $\tau_2$ & Interferon production delays &  \\
    & \hspace{2em} Avian & 21.5 (h) \\
    & \hspace{2em} Seasonal & 23.6 (h) \\
    & \hspace{2em} Pandemic & 21.0 (h) \\
  $\delta$ & Virus-secreting cell decay rate & 0.6 (h$^{-1}$) \\
  $d$ & Chemokine decay rate & 1.386 (h$^{-1}$) \\
  $\tau_1$ & Viral incubation time & 10 (h) \\
  $\tau_3$ & Chemokine production delays & \\
    & \hspace{2em} IP-10 & 8 (h)\\
    & \hspace{2em} RANTES & 16 (h)\\
  \hline
\end{tabular}
\caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.  The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).  Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
\label{tab:dde}
\end{center}
\end{table}
